---
title: "Gaussian VCMoE Simulation Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gaussian VCMoE Simulation Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

This vignette shows the basic steps for a Gaussian
varying-coefficient mixture-of-experts model:

1. install and load `VCMoE`;
2. simulate a small Gaussian example;
3. fit `vcmoe_fit()`;
4. inspect coefficient functions, posterior probabilities, and diagnostics.

The example uses `k = 2`, which is the easiest setting for learning the package
API and checking that the label-alignment diagnostics behave as expected.

## Installation from GitHub

Install the package from GitHub with `remotes`:

```{r install, eval=FALSE}
install.packages("remotes")
remotes::install_github("qc-zhao/Varying-Coefficient-Mixture-of-Experts-Model")
```

Then load the package:

```{r}
library(VCMoE)
```

## Simulate Gaussian data

`simulate_vcmoe_gaussian()` returns a data frame and the known truth used to
generate it. The response is Gaussian, while the component-specific expert
coefficients and gating probabilities vary with the continuous coordinate `u`.

```{r simulate}
sim <- simulate_vcmoe_gaussian(
  n = 240,
  k = 2,
  seed = 1,
  separation = 1.4,
  scenario = "well_separated"
)

head(sim$data)
str(sim$truth, max.level = 1)
```

## Fit the model

The formula has two parts:

- `y ~ z1` is the expert mean model;
- `| x1` is the gating model.

The varying coordinate is supplied separately through `u = "u"`. By default,
`vcmoe_fit()` uses Epanechnikov density weights, a scaled local-linear basis,
and unit scaling of `u`.

```{r fit}
fit <- vcmoe_fit(
  y ~ z1 | x1,
  data = sim$data,
  u = "u",
  family = "gaussian",
  k = 2,
  bandwidth = 0.30,
  u_grid = seq(0.1, 0.9, length.out = 5),
  control = list(maxit = 80, n_starts = 2, seed = 2)
)

fit
```

## Coefficients and predictions

Expert coefficients are returned as an array indexed by grid point, component,
and term. Posterior probabilities are returned with one column per component.

```{r coefficients}
expert_coef <- coef(fit, "expert")
dim(expert_coef)
expert_coef[, , "z1"]
```

```{r predictions}
posterior <- predict(fit, type = "posterior")
head(posterior)
rowSums(head(posterior))

fitted_mean <- predict(fit, type = "mean")
head(fitted_mean)
```

## Diagnostics

Always inspect diagnostics before interpreting coefficient functions. The most
important early checks are grid convergence, label ambiguity, component
proportions, posterior entropy, and effective local sample size.

```{r diagnostics}
diagnostics <- vcmoe_diagnostics(fit)
diagnostics[, c("u", "converged", "ambiguous", "posterior_entropy", "effective_n")]
```

## Plots

`plot_coefficients()` shows the fitted coefficient functions. For a simulated
example, the plot is mainly a quick visual check that the fitted paths are
smooth and labels remain connected across the grid.

```{r coefficient-plot}
plot_coefficients(fit, "expert")
```

`plot_posterior()` summarizes posterior component probabilities over `u`.

```{r posterior-plot}
plot_posterior(fit)
```

## Optional: bandwidth selection

For real analyses, bandwidth should usually be selected rather than fixed by
hand. The cross-validation selector uses held-out predictive log-likelihood and
returns a final refit by default.

```{r bandwidth, eval=FALSE}
selection <- vcmoe_select_bandwidth(
  y ~ z1 | x1,
  data = sim$data,
  u = "u",
  family = "gaussian",
  k = 2,
  bandwidth_grid = c(0.24, 0.30, 0.36),
  folds = 3,
  u_grid = seq(0.1, 0.9, length.out = 5),
  control = list(maxit = 80, n_starts = 2, seed = 3),
  seed = 4
)

selection
selection$best_bandwidth
```
